Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media

### 2015-10-07 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

%\medskip
Under certain conditions the displacement of fluids in porous media is unstable and results in the formation of strange patterns in the form of \textit{fingers}. This phenomenon is referred to as \textit{viscous fingering}. The flow of an injected fluid into a geological formation (porous medium) generates many instabilities due to the heterogeneities of the porous medium. Even though the mechanism of the instabilities may vary (in \textcolor{blue}{immiscible fluids} the most important factor is the surface tension while in the \textcolor{blue}{miscible fluids} it is the diffusion) however it is still a function of many variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover,in most of the cases interfacial tension is directly linked with the mobility ratio -or viscosity ratio- between the displaced and the displacing fluid.

Both the evolution of the computers and the advances in modelling and simulation of flows in porous media may help to visualise and address the importance of this phenomenon which can be found in many petroleum, chemical and environmental related applications.

%\medskip
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}. The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to overcome these deficiencies.

@@

\citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation fields were volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.

@@

\textcolor{blue}{In this paper, the overlapping control volume finite element method (OCVFEM), formulation recently introduced by~\citet{jackson_2013}~\citep[see also][]{gomes_2013,pavlidis_2013} is used to investigate the development of multiphase flows through heterogeneous porous media. In the OCVFEM, a dual pressure-velocity representation in CV and FEM spaces  are embedded into a family of triangular/tetrahedral element pairs, P$_{m}$DG-P$_{n}$~\citep{cotter_2009b} and P$_{m}$DG-P$_{n}DG$~\citep{xie_2014}. In these element types, the dual velocity and pressure fields are represented by $m^{\rm th}$- (discontinuous) and $({n})^{\rm th}$-order (continuous/discontinuous) polynomials, respectively.
Moreover, velocity field has dual mapping in FEM and CV spaces. This approach can achieve an exact representation of the Darcy force balance equations.\textcolor{red}{The main objective of this paper is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media}. 
In addition, the study focused in flow instabilities determined by the viscosity ratio. In the simulations presented in this paper the element pairs as described above and the overlapping control volume finite element method (CVFEM) model were used, in which the dual velocity-pressure fields are embedded in FEM space with simultaneous projection into CV space. Scalar fields (e.g., saturation, concentration, density, etc) are represented in the control volume space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section.}

>>>>

%\medskip
Under certain conditions the displacement of fluids in porous media is unstable \red{(what sort of conditions?)} and results in the formation of strange patterns \red{(sentence is not very technical} in the form of \textit{fingers}. This phenomenon is referred to as \textit{viscous fingering}. The flow of an injected fluid into a geological formation (porous medium) generates instabilities due to the heterogeneities of the porous medium \red{(Is this correct? Is it your opnion? If so built up the facts to substantiate your conclusions. If it is not your opinion then give reference(s))}. Even though the mechanism of the instabilities may vary (in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)}), however it is still a function of several variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover, in most cases interfacial tension is directly linked with the mobility ratio (or viscosity ratio) between the displaced and the displacing fluid.

Both the evolution of the computers and the advances in modelling and simulation of flows in porous media may help to visualise and address the importance of this phenomenon which can be found in many petroleum, chemical and environmental related applications. \red{(sentence is not written in a technical format, also what is its objective?)}

%\medskip
\begin{comment}
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}. The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to overcome these deficiencies.

@@

\citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation fields were volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.
\end{comment}

@@

In this paper, the overlapping control volume finite element method (OCVFEM), formulation recently introduced by~\citet{jackson_2013}~\citep[see also][]{gomes_2013,pavlidis_2013} is used to investigate the development of multiphase flows through heterogeneous porous media \red{(what is the objective of this paper? To study multiphase porous media flows? Or to study a particular physical aspect of multiphase flows in porous media? You need to be precise here! )}. In the OCVFEM, a dual pressure-velocity representation in CV and FEM spaces  are embedded into a family of triangular/tetrahedral element pairs, P$_{m}$DG-P$_{n}$~\citep{cotter_2009b} and P$_{m}$DG-P$_{n}DG$~\citep{xie_2014}. In these element types, the dual velocity and pressure fields are represented by $m^{\rm th}$- (discontinuous) and $({n})^{\rm th}$-order (continuous/discontinuous) polynomials, respectively.
Moreover, velocity field has dual mapping in FEM and CV spaces. This approach can achieve an exact representation of the Darcy force balance equations.\textcolor{red}{The main objective of this paper is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media}. 
In addition, the study focused in flow instabilities determined by the viscosity ratio. In the simulations presented in this paper the element pairs as described above and the overlapping control volume finite element method (CVFEM) model were used, in which the dual velocity-pressure fields are embedded in FEM space with simultaneous projection into CV space. Scalar fields (e.g., saturation, concentration, density, etc) are represented in the control volume space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section.
